---
title: "Outlier Protein and Phosphosite Target Identifier"
output: pdf_document
# bibliography: bibliography.bib
vignette: |
  %\VignetteIndexEntry{X. Outlier Protein and Phosphosite Target Identifier} %\VignetteEngine{knitr::rmarkdown}
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300)
knitr::opts_chunk$set(cache=FALSE)
```
  
```{r, echo = FALSE, hide=TRUE, message=FALSE, warning=FALSE}
devtools::load_all(".")
```

# Analyze proteomics data of a single cohort
You can easily analyze outlying (dysregulated) markers for each sample in a cohort. Lets generate a toy proteomics data for a cohort of 30 tumor samples, each quantifying 100 proteins.
```{r, eval = TRUE}
set.seed(1)
cohort1.proteome = as.data.frame(matrix(abs(rnorm(100*30)), 100, 30)) 
rownames(cohort1.proteome) = paste('marker', 1:100, sep = '')
colnames(cohort1.proteome) = paste('cohort1.sample', 1:30, sep = '')
```
Outlier analysis is run by the `oppti` function:
```{r, eval = TRUE}
library(oppti)
result = oppti(cohort1.proteome)
```

The outlier scores of each marker in each sample are then returned in the first element of the result:
```{r, eval = TRUE} 
cohort1.outlier.scores = result[[1]] 
```
```{r, eval = TRUE, echo = FALSE, size = 8}
knitr::kable(cohort1.outlier.scores[1:10,1:4], digits = 2, 
             caption = "Example matrix of outlier scores, displayed for the first 10 proteins (rows) and the first 4 samples (columns)",
             row.names = TRUE)
```

In this toy example, marker5 has a (somewhat) elevated outlier score in sample3, suggesting a protruding expression in the tumor tissue of sample3 relative to a normal state (i.e., the consensus co-expression network inferred for marker5). Note that a negative sign in the outlier score indicates a negative dysregulation event, i.e., relatively "lower" protein expression is expected in the observed tumor state compared to the normal state. The landscape of these aberrant expressions analyzed for a cohort of individuals may serve for the discovery of personalized actionable targets.

# Analyze proteomics data of multiple cohorts
For pan-cancer analyses, the normalized proteomics data from different cohorts can be supplied to `oppti` in a list object. Lets generate another toy proteomics data for a separate cohort of 20 tumor samples, each quantifying 80 proteins (say, 50 of which are overlapping with those quantified in the first cohort).
```{r, eval = TRUE}
cohort2.proteome = as.data.frame(matrix(abs(rnorm(80*20)), 80, 20)) 
rownames(cohort2.proteome) = paste('marker', 51:130, sep = '')
colnames(cohort2.proteome) = paste('cohort2.sample', 31:50, sep = '')
```

To run `oppti` for both cohorts, the data are simply fed in a single list object:
```{r, eval = TRUE}
result = oppti(list(cohort1.proteome,cohort2.proteome))
```

Again, the outlier scores of each marker in each sample are returned in the first element of the result. 
```{r, eval = TRUE} 
outlier.scores = result[[1]]
```

However, this object is a list of 2 elements per se, corresponding to two cohorts. To obtain the outlier scores of the first cohort:
```{r, eval = TRUE} 
cohort1.outlier.scores = outlier.scores[[1]]
```
```{r, eval = TRUE, echo = FALSE, size = 8}
knitr::kable(cohort1.outlier.scores[1:10,1:4], digits = 2, 
             caption = "Example outlier scores in cohort1",
             row.names = TRUE)
```

Similarly, for the second cohort the outlier scores are obtained by:
```{r, eval = TRUE} 
cohort2.outlier.scores = outlier.scores[[2]]
```
```{r, eval = TRUE, echo = FALSE, size = 8}
knitr::kable(cohort2.outlier.scores[1:10,1:4], digits = 2, 
             caption = "Example outlier scores in cohort2",
             row.names = TRUE)
```

